Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2016-02-05 WSSSPE3/report/wg_appendix_eng_design.tex
small updates

This group is comprised  of members from different backgrounds, including quantum
chemistry, epidemiology, microscopy, computer science, and software engineering.
Each participant was invited to give their perspective on the topic area and
what they thought were the crucial points for discussion. There was a general
consensus that there was a need for relating principles to practice for the
computational science and engineering community. Furthermore, various members of
the group expressed their interest in tools and best practices for facilitating
the maintenance and evolution of scientific software systems. It was agreed to
identify principles from software engineering and from sustainability design
and, based on those lists, discuss what each of those would mean applied to
specific example systems from the expert domains of some of the group members.
The group identified a number of software engineering principles drawn from the
Software Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.

Software design principles include abstraction; coupling and cohesion;
decomposition and modularization; encapsulation and information hiding;
separation of interface and implementation; sufficiency completeness and
primitiveness; and separation of concerns. Similarly, user interface design
principles include learnability; user familiarity; consistency; minimal
surprise; recoverability; user guidance; and user diversity. The sustainability
design principles were drawn from the Karlskrona Manifesto on Sustainability
Design~\cite{Becker:2014}. The manifesto states that sustainability is systemic,
multidimensional, and interdisciplinary; transcends the system's purpose; applies to
both a system and its wider contexts; requires action on multiple levels;
requires multiple timescales; changing design to take into account long-term
effects does not automatically imply sacrifices; system visibility is a
precondition for and enabler of sustainability design. A number of sustainable
software engineering principles proposed by Tate~\cite{tate:2005} were also
considered including: continual refinement of product and project practices; a
working product at all times; continual emphasis on design; and value defect
prevention over defect detection.

@@

In order to achieve the following three goals: (1) a systematic analysis of a number of example systems from
different scientific domains with regards to the identified principles, (2) the
identification of the commonalities and gaps in applying the principles to different
scientific systems, and (3) a proposal of a set of guidelines on the
principles, the next steps were discussed in the following subsections.

>>>>

This group was comprised  of members from different backgrounds, including quantum
chemistry, epidemiology, microscopy, computer science, and software engineering.
Each participant was invited to give their perspective on the topic area and
what they thought were the crucial points for discussion. There was a general
consensus that there was a need for relating principles to practice for the
computational science and engineering community. Furthermore, various members of
the group expressed their interest in tools and best practices for facilitating
the maintenance and evolution of scientific software systems. It was agreed to
identify principles from software engineering and from sustainability design
and, based on those lists, discuss what each of those would mean applied to
specific example systems from the expert domains of some of the group members.
The group identified a number of software engineering principles drawn from the
Software Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.

Software design principles include abstraction, coupling and cohesion,
decomposition and modularization, encapsulation and information hiding,
separation of interface and implementation, sufficiency completeness and
primitiveness, and separation of concerns. Similarly, user interface design
principles include learnability, user familiarity, consistency, minimal
surprise, recoverability, user guidance, and user diversity. The sustainability
design principles were drawn from the Karlskrona Manifesto on Sustainability
Design~\cite{Becker:2014}. The manifesto states that sustainability is systemic,
multidimensional, and interdisciplinary; transcends the system's purpose; applies to
both a system and its wider contexts; requires action on multiple levels;
requires multiple timescales; changing design to take into account long-term
effects does not automatically imply sacrifices; system visibility is a
precondition for and enabler of sustainability design. A number of sustainable
software engineering principles proposed by Tate~\cite{tate:2005} were also
considered including: continual refinement of product and project practices; a
working product at all times; continual emphasis on design; and value defect
prevention over defect detection.

@@

In order to achieve the following three goals: (1) a systematic analysis of a number of example systems from
different scientific domains with regards to the identified principles, (2) the
identification of the commonalities and gaps in applying the principles to different
scientific systems, and (3) a proposal of a set of guidelines on the
principles, the following next steps were discussed.

### 2016-02-05 WSSSPE3/report/wg_main_eng_design.tex
small updates

Software engineering principles form the foundation of methods, techniques,
methodologies, and tools.  Consisting of members from different
backgrounds, including quantum chemistry, epidemiology, computer science,
software engineering, and microscopy, this group discussed the principles of software
engineering design for sustainable software (starting with principles from the
Karlskrona Manifesto on Sustainability Design~\cite{Becker:2015},
Tate~\cite{tate:2005}, and the Software Engineering Body of Knowledge
(SWEBOK)~\cite{swebokv3}) and their applications in various domains including
quantum chemistry and epidemiology. The group examined the principles and took a
retrospective analysis of what the developers did in practice against how the
principles could have made a difference, and asked, what do the principles mean
for computational scientific and engineering software, and how do the principles
relate to non-functional requirements? It appeared that the sustainable software
engineering principles should be mapped to two core quality attributes that
underpin technically sustainable software: \emph{extensibility}, the software's ability
to be extended and the effort level required to implement the extension; and
\emph{maintainability}: the effort required to locate and fix an error in operational
software.

>>>>

Software engineering principles form the foundation of methods, techniques,
methodologies, and tools.  Consisting of members from different
backgrounds, including quantum chemistry, epidemiology, computer science,
software engineering, and microscopy, this group discussed the principles of software
engineering design for sustainable software (starting with principles from the
Karlskrona Manifesto on Sustainability Design~\cite{Becker:2015},
Tate~\cite{tate:2005}, and the Software Engineering Body of Knowledge
(SWEBOK)~\cite{swebokv3}) and their application in various domains including
quantum chemistry and epidemiology. The group examined the principles and took a
retrospective analysis of what the developers did in practice against how the
principles could have made a difference, and asked, what do the principles mean
for computational scientific and engineering software, and how do the principles
relate to non-functional requirements? It appeared that the sustainable software
engineering principles should be mapped to two core quality attributes that
underpin technically sustainable software: \emph{extensibility}, the software's ability
to be extended and the effort level required to implement the extension; and
\emph{maintainability}: the effort required to locate and fix an error in operational
software.
